The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2013-10-28 paper/ms.tex
edits in the results section. not done yet.

\begin{abstract}
We present results of radiative transfer calculations to measure
the impact of gas bulk rotation on the morphology of the Lyman
$\alpha$ line in galaxies. We model a galaxy as a sphere with an
homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with
maximum velocities in the range $0-300$ \kms and neutral
hydrogen optical depths in the range $\tau_{\rm
H}=10^{5}-10^{7}$. We also consider two kinds of spatial
distribution for the radiation sources with respect to the sphere:
central and homogeneous. We find that the line width and the amount of
Lyman $\alpha$ photons escaping in the line center increase with
the rotational velocity. In the case of homogeneously distributed
sources, for large rotational velocities the line transforms from a
double peak to a single peak a the line center. Under the same
conditions the escape fraction increases by a factor of $\sim 2$. For
radiation sources located off-center, the line morphology presents a
range of single, double and triple peaked lines. We show how these results are
useful to interpret recent spectroscopic results of distant $z\sim 2-3$ star
forming galaxies.
\end{abstract}

@@

The central result of this paper is summarized in Figure
\ref{fig:differentvelocities}. It shows the considerable
impact of rotation on the morphology of the emergent \ly line. Both
panels in the Figure focus on the results for $\tau_{H}=10^{7}$,
showing that the influence of rotation is present both when the
photons are either homogeneously or centrally initialized over the gas
volume.

In the following subsections we characterize the line morphology by
the half-width at half intensity and the peak maxima. In order to
interpret the morphological changes in the line we also report the
median number of scatter for each \ly photon in the
simulation. For the models where dust is included we measure the bulk
escape fraction as a function of rotational velocity. Finally, we make
an estimate of the anisotriopic emission of the models in comparison
with static spheres.

@@

To quantify the line broadening we use a modified version of the full
width at half maximum (FWHM). We measure it only for half of
the line, $W_{1/2}$. This definition allows us to quantify the line
width and see the transition from double to single peak emission.  In the
case of double peaked emission $W_{1/2}$ corresponds to the width of a
single peak, while in the extreme case of high rotational velocities,
when the double peak is erased, it simply correspond to half FWHM.

Figure \ref{fig:widthvsvelocity} shows how $W_{1/2}$ increases with
rotational velocity. Continuous (dashed) lines connect the results for
homogeneous (central) source distribution. For the temperature
$T=10^4$K used in our radiative transfer calculations the thermal
velocity is $v_{th}=12.8$\kms. For a model with $\tau_{H}$ it means
that the half-width can increase up to $350$\kms (at $V_{\rm
  max}=300$\kms) compared to a half-width of $150$\kms in the static
case.

Figure \ref{fig:maximumsvsvelocity} shows the position for the peak
maxima as a function of the rotational velocity $V_{\rm max}$. This
Figure clearly shows that in the case of central distributed
sources there is barely any change with rotational velocity in the
range of explored parameters. However, in the case of
homogeneously emitted sources the maxima position remain close to
constant until beyond some velocity threshold the line becomes single
peaked with $x_{\rm max}=0$ \kms.

The transition to a single peak line seems to occur for systems
where it becomes easier for the bulk of the photons to escape with the lowest
number of scatterings possible, allowing them not to move very far
from the center of the line. This might explain how the single peak stage
is easily achieved in the homogeneous source distribution. In this
case there is a fraction of the photons that are inside a photosphere
region with $\tau_{H,r}\ll \tau_{H}$ where $\tau_{H,r}$ is the optical
depth from the radius of emission to the sphere's surface. This
conditions allows the photons to escape with much less scatterings
compared to the photons emitted at the very center of the sphere. In
turn, it gives the photons less scatterings to be placed far from the
line center.

@@

For the central emission the transition to a single peak is never
completed in the range of parameters explored here. The non appearance
of a single peak phase can be in part explained to the absence of a
photosphere, as is the case in the homogeneous
distribution. Nevertheless, there is a rise in the intensity at the
line center as the rotational velocity increases. This hints that the
encounters with a non static medium are inefficient in changing a
photons' frequency outside the line center.

In the next section we quantify the number of scatterings and explore
to what extent this is related to the emergence of single peak.

@@

\begin{figure}
    \includegraphics[width=0.45\textwidth]{NscattvsVmax.png}
\caption{Logarithm of the average number of scatterings as function of
  the velocity. Continuous/dashed lines represent an
  homogeneous/central distribution of sources. \label{fig:Nscatt}}    
\end{figure}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{2dHistogram0t5HOM.png}
  \includegraphics[width=0.45\textwidth]{2dHistogram300t5HOM.png}\\
  \includegraphics[width=0.45\textwidth]{2dHistogram0t5.png}
  \includegraphics[width=0.45\textwidth]{2dHistogram300t5.png}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) pannels
      show the homogeneous (central) source distribution. Left
      corresponds to the static case and the right
      $V_{max}=300km/s$. The colour scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

The number of times that a \ly photon is absorbed and re-emitted is
affects the final frequency that it will have after escaping the
galaxy. In the case of static gas geometries, a large value of the
optical depth correlates with a high number of
scatterings. This increases the probability of finding
a \ly photon far from the center of the line. In summary, the peak
maxima shift away form the line center as the amount of neutral
hydrogen increases.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central distributions the average number of
scatterings has a modes change for increasing rotational
velocities, $\langle N_{\rm scatt}\rangle$ changes less than $0.5\%$
for different velocities. In the same setup the average
number of scatterings is proportional to the optical depth, as
expected in analogy from the analytic result for the homogeneous
infinite-slab $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}. In static spheres 
with centrally distributed sources we find $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depths
$\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$

Figure \ref{fig:Nscatt} shows that for the homogeneous distribution
there is a clear decrease of $\langle N_{\rm  scatt}\rangle$ as the
$V_{\rm max}$ increases. This effect more pronounced for the lower
values of the optical depth. For $\tau_{\rm H}=10^5$ the average
number of scatterings decreases by $61\%$ at $V_{\rm max}=300\kms$ in
comparison to the static case.

The analytic expectation for the slab with homogeneously emitted
sources is $\langle N_{\rm  scatt}\rangle=1.16\tau_{\rm   H}$
\citep{Harrington73}, a factor of $0.72$ lower than the case of the
centrally emitted photons. In our case we find that for the static
, while for homogeneously distributed source $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this represents 
  a factor of $(0.66, 0.59, 0.51)$ lower than the centrally emitted photons.

In order to gain a deeper understanding of these results we prepare 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show
the results of such histogram in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$. This Figure clearly support our hypothesis about
the photosphere in the homogeneous distribution, in this case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings, explaining the origin of a single central
peak. However, for a central distribution the situation is
different. In this case the number of scatterings remains high, on the
order of the optical depth, but the two peaks do get closer to each
other. In this case the physical picture is that each scattering, due
to the bulk velocity of the gas, is inefficient in driving the photon
outside the line center.

@@

We also study a dusty cloud configuration to measure the effect of
rotation on the escape fraction. We expect that the modified number of
scatterings to be reflected in this amount of photons absorbed by
dust. Following this line of thought,  we do not expect any change in
for a dusty cloud with central source of radiation given that the
number of scatterings remains close to constant. On the other hand, in
the case of an homogeneous radiation source the number of scatterings
drops as $V_{\rm max}$ increases, which might be reflected as an
increasing escape fraction.

Figure \ref{fig:efvsv} shows the dependence of the escape fraction as
a function of the maximum rotational velocity, confirming that our
intuitions in this respect is correct. For the central source
distribution the escape fraction barely shows any change, while for
the homogeneous case there is a clear rise in the escape fraction for
high rotational velocities.  Rotation has a higher relative impact in
the models with low optical depth $\tau_{\rm H}=10^5$,$10^{6}$, where
it can raise from $(0.26, 0.28)$ respectively in the static case up to $(0.37, 0.39)$ at $V_{\rm
  rot}=300$\kms.

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytical solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends solely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} correspond to the cases of different
velocities. We observe that the escape fraction is higher
by factors of $2-10$  than the expected values for the slab
configuration. We also see that for the lower value $\tau_{\rm
  H}=10^{5}$ the escape fraction does not increase with respect to the
solution for $\tau_{\rm H}=10^{6}$ as expected, however we not that we
are in a regime where the condition for the analytic expectations 
($a\tau_{h}\gg 1$) does not hold.

@@

As we know there is unlikely to find galaxies with radiation sources 
distributed homogeneously. Most of them are in a clumpy distribution 
(Laursen et al 2013*) which affected the resulting spectra. In order 
to study an inhomogeneous distribution we set up a model in which we 
select certain photons that are placed in a specific place but that 
are not symmetrically distributed Fig. \ref{fig:OCspheres} shows the 
distributions we set up, basically we take photons located in a sphere
 of $0.5 R$ where $R$ is the radii of the homogeneous sphere and 
we choose two spheres in each side of the $\hat{\textbf{x}}$ axis,
two at $\mp 0.25 R$ and the other two at $\mp 0.5 R$.

\begin{figure}
\includegraphics[scale=0.3]{OF_spheres.png}
\caption{The two solid circles show up the photons we select in order to make the anisotropic emission
		rotation is defined to be $\mathrm{z}$-axis. we also select photons in the negative side of 				        $\mathrm{x}$ (dashed circles) in order to see the complete effect of rotation. 
   \label{fig:OCspheres}} 
\end{figure}

We study the deviations of the received flux on the surface of the
sphere. We follow \cite{Zheng2013} to estimate the flux seen by an
observer located at an angle $\Theta$ (distant observer angle)
normalized to the isotropic flux:

@@

For the central and homogeneous source distributions we find
that for observers located at  $\theta = pi/2$ the
angular dependency represented by $\mu$ doesn't show a considerable
variation in particular the observer would infer luminosities 
different from the isotropical value only at levels $< 3\%$.

In the case of the Off-center model with the set up showed in 
Fig.\ref{fig:OCspheres} where the galaxy is rotating around the  
$\hat{\textbf{z}}$ axis and the observer is also located at 
$\mathrm{\theta = \pi/2}$, For the special purpose to get a
better sample we used $\tau_{\gamma} = 10^{6}$. The first result 
we get is that the angular dependency is almost the same for 
all the models with different $\tau_{H}$ and $V_{max}$ for 
$R = 0.25$. While for spheres located at and $R = 0.5$ show 
variation in the flux in values $< 14 \%$ in particular at 
$\mu = 0$ the flux in stronger. This results are shown 
in Fig. \ref{fig:OCflux}.

>>>>

\begin{abstract}
We present results of radiative transfer calculations to measure
the impact of gas bulk rotation on the morphology of the Lyman
$\alpha$ line in galaxies. We model a galaxy as a sphere with an
homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with
maximum velocities in the range $0-300$ \kms and neutral
hydrogen optical depths in the range $\tau_{\rm
H}=10^{5}-10^{7}$. We also consider two kinds of spatial
distribution for the radiation sources with respect to the sphere:
central and homogeneous. We find that the line width and the intensity
at the line's center increase with rotational velocity. In the case of homogeneously distributed
sources, for large rotational velocities the line transforms from a
double peak to a single peak a the line center. Under the same
conditions the escape fraction increases $\sim 30\%$. For
radiation sources located off-center, the line morphology presents a
range of single, double and triple peaked lines. We show how these results are
useful to interpret recent spectroscopic results of distant $z\sim 2-3$ star
forming galaxies.
\end{abstract}

@@

The central result of this paper is summarized in Figure
\ref{fig:differentvelocities}. It shows the evident
impact of rotation on the morphology of the emergent \ly line. Both
panels focus on the results for $\tau_{H}=10^{7}$, showing that the
influence of rotation is present both when the photons are either
homogeneously or centrally initialized over the gas volume.

In the next subsections we characterize the line morphology by
the half-width at half intensity and the peak maxima. In order to
interpret the morphological changes in the line we also report the
median number of scatter for each \ly photon in the
simulation. For the models where dust is included we measure the 
escape fraction as a function of rotational velocity. Finally, we make
an estimate of the anisotriopic emission of the models in comparison
with static spheres.

@@

To quantify the line broadening we use the width at half maximum
(FWHM), $W$. Figure \ref{fig:widthvsvelocity} shows how $W$ increases
with rotational velocity. Continuous (dashed) lines connect the
results for homogeneous (central) source distribution. For the
temperature $T=10^4$K used in our radiative transfer calculations the
thermal velocity is $v_{th}=12.8$\kms. For a model with $\tau_{H}$ it
means that the FWHM increase up to $350$\kms (at $V_{\rm
  max}=300$\kms) compared to $150$\kms in the static case.

Figure \ref{fig:maximumsvsvelocity} shows the position for the peak
maxima as a function of the rotational velocity $V_{\rm max}$. This
Figure clearly shows that in the case of central distributed
sources there is barely any change with rotational velocity in the
range of explored parameters. However, in the case of
homogeneously emitted sources the maxima position remain close to
constant until beyond some velocity threshold the line becomes single
peaked with $x_{\rm m}=0$ \kms.

The transition to a single peaked line seems to occur in systems
where it is easier for the bulk of the photons to escape with the lowest
number of scatterings possible, allowing them not to move very far
from the center of the line. This might explain how the single peak stage
is easily achieved in the homogeneous source distribution. In this
case there is a fraction of the photons that are inside a photosphere
region with $\tau_{H,r}\ll \tau_{H}$ where $\tau_{H,r}$ is the optical
depth from the radius of emission to the sphere's surface. This
conditions allows the photons to escape with much less scatterings
compared to the photons emitted at the very center of the sphere. In
turn, it gives the photons less scatterings to be placed far from the
line center.

@@

For the central emission the transition to a single peak is never
completed in the range of explored parameters. The non appearance
of a single peak phase can be in part explained to the absence of a
photosphere, as is the case in the homogeneous
distribution. Nevertheless, there is a rise in the intensity at the
line center as the rotational velocity increases. This hints that the
encounters with a non static medium are inefficient in pushing photons
outside the line's center.

In the next sub-section we quantify the number of scatterings and explore
to what extent this is related to the emergence of single peaked emission.

@@

\begin{figure}
    \includegraphics[width=0.45\textwidth]{NscattvsVmax.png}
\caption{Logarithm of the average number of scatterings as function of
  rotational velocity velocity. Continuous (dashed) lines represent an
  homogeneous (central) distribution of sources. \label{fig:Nscatt}}    
\end{figure}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{2dHistogram0t5HOM.png}
  \includegraphics[width=0.45\textwidth]{2dHistogram300t5HOM.png}\\
  \includegraphics[width=0.45\textwidth]{2dHistogram0t5.png}
  \includegraphics[width=0.45\textwidth]{2dHistogram300t5.png}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left
      corresponds to the static case and the right
      $V_{max}=300km/s$. The colour scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

The number of \ly photon scatterings affects its final frequency after
escaping the galaxy. In static gas geometries, a large value of the
optical depth correlates with a high number of scatterings, increasing
the probability of finding a \ly photon far from the center of the
line. In these cases the peak maxima shift away form the line center as
the amount of neutral hydrogen increases.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central distributions the average number of
scatterings has a modest change for increasing rotational
velocities, $\langle N_{\rm scatt}\rangle$ changes less than $0.5\%$
for different velocities. In the same setup the average
number of scatterings is proportional to the optical depth, as
expected in analogy from the analytic result for the homogeneous
infinite-slab $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}. In the spheres with centrally
distributed sources we find $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depths
$\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$

Figure \ref{fig:Nscatt} shows that for the homogeneous distribution
there is a clear decrease of $\langle N_{\rm  scatt}\rangle$ as the
$V_{\rm max}$ increases. This effect is more pronounced for lower
values of the optical depth. For $\tau_{\rm H}=10^5$ the average
number of scatterings decreases by $61\%$ at $V_{\rm max}=300\kms$ in
comparison to the static case.

The analytic expectation for the slab with homogeneously distributed
sources is $\langle N_{\rm  scatt}\rangle=1.16\tau_{\rm   H}$
\citep{Harrington73}, a factor of $0.72$ lower than the case of the
centrally emitted photons. In our case we find that for the static
, while for homogeneously distributed source $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this represents 
  factors of $(0.66, 0.59, 0.55)$ lower than the centrally emitted
  photons.

In order to gain a deeper understanding of these results we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show
such histogram in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$.

Figure \ref{fig:Nscatt2D} supports our hypothesis about
the photosphere in the homogeneous distribution. In that case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This explains the origin of a single central
peak. However, for a central distribution the situation is
different. In this case the number of scatterings remains high, on the
order of the optical depth, but the two peaks do get closer to each
other. In this case the physical picture is that each scattering, due
to the bulk velocity of the gas, is inefficient in driving the photon
outside the line center.

@@

The average number of scatterings is also reflected in the amount of
photons absorbed by dust. Under this considerations we do not expect
any change in the escape fraction for a dusty cloud with central
source of radiation given that the number of scatterings remains close
to constant. On the other hand, in the case of an homogeneous
radiation source the number of scatterings drops as $V_{\rm max}$
increases, which has to be reflected as an increasing escape fraction.

Figure \ref{fig:efvsv} shows the dependence of the escape fraction as
a function of the maximum rotational velocity, confirming our expectations. For the central source
distribution the escape fraction barely shows any change, while for
the homogeneous case there is a clear rise in the escape fraction for
high rotational velocities.  Rotation has a higher relative impact in
the models with low optical depth $\tau_{\rm H}=10^5$,$10^{6}$, where
it can rise from $(0.26, 0.28)$ ($V_{\rm}=0$\kms) up to $(0.37, 0.39)$ ($V_{\rm
  rot}=300$ \kms).

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytical solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends solely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} correspond to the cases of different
velocities. First of all we note that the escape fraction does not
increase from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, this is
explained because we are in a regime where the condition for the
analytic expectations  ($a\tau_{h}\gg 1$) does not hold. This figure
also shows that the escape fraction is higher by factors of $2-10$
than the expectation for a slab with centrally distributed
sources. Furthermore, the decrease of the escape fraction with
increasing optical depth is also slower.

@@

\begin{figure}
  \includegraphics[scale=0.3]{OF_spheres.png}
  \caption{The two solid circles show the two different spherical off-centered emitting
    regions. The angular velocity vector is defined to be along the
    $\hat{\i}$ direction and the observer is along the $\hat{\j}$ direction (where
    the reader is). 
    \label{fig:OCspheres}} 
\end{figure}

The model with homogeneous radiation sources allows to study the
effect of radiation sources distributed with a different symmetry than
the gas. We select photons emitted in spheres that are off-centered
with respect to the gas distribution.

Figure \ref{fig:OCspheres} shows the 
two spherical regions we define. Each has a radius of $0.5$R and can
have to possible centers positions $C$ along the $\hat{\i}$ direction:
$\pm C\hat{\i}$. We also place an observer along the $\hat{\j}$
direction, such that the emitting sphere centered on $+C\hat{\i}$ moves away
from the observer and the sphere centered on $+C\hat{\i}$ moves towards
the observer. We consider two values for $C=(0.25,0.50)R$.

We study first the deviations of the received flux on the surface of the
sphere. We follow \cite{Zheng2013} to estimate the flux seen by an
observer located at an angle $\Theta$ (distant observer angle)
normalized to the isotropic flux:

@@

For the central and homogeneous source distributions we find that the
level of isotropy angular dependency represented by $\mu$ doesn't show
a considerable variation in particular the observer would infer
luminosities  different from the isotropical value only at levels $<
3\%$. This values were obtained for a run with $V_{\rm max}=XX$ and
$\tau_{\rm H}=XXX$ and $N_{\gamma}=10^{6}$ photons.

%%%%%%%
